Counterfactual Training: Teaching Models Plausible and Actionable Explanations

### 2025-02-24 paper/preprint/paper.tex
First review commit
Fixes some typos and minor inconsistencies to initialize review

\usepackage{lineno}
\linenumbers
\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Counterfactual Training (A Preprint) }
\title{Counterfactual Training: Teaching Models Plausible and Actionable
Explanations}
\def\asep{\\\\\\ } % default: all authors on same column
\def\asep{\And }
\author{\textbf{Patrick
Altmeyer}~\orcidlink{0000-0003-4726-8613}\\Faculty of Electrical
Engineering, Mathematics and Computer Science\\Delft University of
Technology\\\\\href{mailto:p.altmeyer@tudelft.nl}{p.altmeyer@tudelft.nl}\asep\textbf{Aleksander
Buszydlik}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\\asep\textbf{Arie van
Deursen}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\\asep\textbf{Cynthia C. S.
Liem}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\}
\date{}
\begin{document}
\maketitle
\begin{abstract}
Counterfactual Explanations have emerged as a popular tool to explain
predictions made by opaque machine learning models: they explain how
factual inputs need to change in order for some fitted model to produce
some desired output. Much existing research has focused on identifying
explanations that are not only valid but also deemed plausible and
desirable with respect to the underlying data and stakeholder
requirements. Recent work has shown that under this premise, the task of
learning plausible explanations is effectively reassigned from the model
itself to the (post-hoc) counterfactual explainer. Building on that
work, we propose a novel model objective that leverages counterfactuals
during the training phase (ad-hoc) in order to minimize the divergence
between learned representations and plausible explanations. Through
extensive experiments, we demonstrate that our proposed methodology
facilitates training models that inherently deliver plausible
explanations while maintaining high predictive performance.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
Counterfactual Explanations \sep Explainable AI \sep 
Representation Learning

@@

Part of the problem is that high degrees of freedom provide room for
many solutions that are locally optimal with respect to narrow
objectives (\citeproc{ref-wilson2020case}{Wilson 2020})\footnote{For
  clarity: we follow standard ML convention in using ``degrees of
  freedom'' to refer to the number of parameters estimated from data.}.
Based purely on predictive performance, these solutions may seem to
provide compelling explanations for the data, when in fact they are
based on purely associative, semantically meaningless patterns. This
poses two related challenges: firstly, it makes these models inherently
opaque, since humans cannot simply interpret what type of explanation
the complex learned representations correspond to; secondly, even if we
could resolve the first challenge, it is not obvious how to mitigate
models from learning representations that correspond to meaningless and
implausible explanations.

@@

Counterfactual training combines ideas from adversarial training,
energy-based modelling and counterfactuals explanations with the
explicit objective of aligning representations with plausible
explanations that comply with user requirements. In the context of CE,
plausibility has broadly been defined as the degree to which
counterfactuals comply with the underlying data generating process
(\citeproc{ref-poyiadzi2020face}{Poyiadzi et al. 2020};
\citeproc{ref-guidotti2022counterfactual}{Guidotti 2022};
\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}). Plausibility
is a necessary but insufficient condition for using CE to provide
algorithmic recourse (AR) to individuals affected by opaque models in
practice. This is because for recourse recommendations to be
\textbf{actionable}, they need to not only result in plausible
counterfactuals but also be attainable. A plausible CE for a rejected
20-year-old loan applicant, for example, might reveal that their
application would have been accepted, if only they were 20 years older.
Ignoring all other features, this complies with the definition of
plausibility if 40-year-old individuals are in fact more credit-worthy
on average than young adults. But of course this CE does not qualify for
providing actionable recourse to the applicant since \emph{age} is not a
mutable feature. For our intents and purposes, counterfactual training
aims at improving model explainability by aligning models with
counterfactuals that meet both desiderata, plausibility and
actionability. Formally, we define explainability as follows:

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  (Plausibility)
  \(\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1\)
  where \(A\) is some small region around \(\mathbf{x}^{\prime}\).
\item
  (Actionability) Permutations \(\Delta\) are subject to actionability
  constraints.
\end{enumerate}

@@

Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote a counterfactual
explanation generated through gradient descent over \(T\) iterations as
initially proposed by Wachter, Mittelstadt, and Russell
(\citeproc{ref-wachter2017counterfactual}{2017}). For our purposes, we
let \(T\) vary and consider the counterfactual search as converged as
soon as the predicted probability for the target class has reached a
pre-determined threshold, \(\tau\):
\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\)\footnote{For
  detailed background information on gradient-based counterfactual
  search and convergence see Section~\ref{sec-app-ce}.}.

@@

Intuitively, the gradient of Equation~\ref{eq-div} decreases the energy
of observed training samples (positive samples) while at same time
increasing the energy of counterfactuals (negative samples)
(\citeproc{ref-du2019implicit}{Du and Mordatch 2020}). As the generated
counterfactuals get more plausible (Definition~\ref{def-explainability})
over the cause of training, these two opposing effects gradually balance
each out (\citeproc{ref-lippe2024uvadlc}{Lippe 2024}).

@@

Under this choice, we consider nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\) as adversarial examples as long as the
magnitude of the perturbation to any individual feature is at most
\(\varepsilon\). This is closely aligned with Szegedy et al.
(\citeproc{ref-szegedy2013intriguing}{2013}), who define an adversarial
attack as an ``imperceptible non-random perturbation''. Thus, we choose
to work with a different distinction between CE and AE than Freiesleben
(\citeproc{ref-freiesleben2022intriguing}{2022}), who considers
misclassification as the key distinguishing feature of AE. One of the
key observations in this work is that we can leverage counterfactual
explanations during training and get adversarial examples, essentially
for free.

>>>>

\usepackage{lineno}
\linenumbers
\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Counterfactual Training (A Preprint) }
\title{Counterfactual Training: Teaching Models Plausible and Actionable
Explanations}
\def\asep{\\\\\\ } % default: all authors on same column
\def\asep{\And }
\author{\textbf{Patrick
Altmeyer}~\orcidlink{0000-0003-4726-8613}\\Faculty of Electrical
Engineering, Mathematics and Computer Science\\Delft University of
Technology\\\\\href{mailto:p.altmeyer@tudelft.nl}{p.altmeyer@tudelft.nl}\asep\textbf{Aleksander
Buszydlik}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\\asep\textbf{Arie van
Deursen}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\\asep\textbf{Cynthia C. S.
Liem}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\}
\date{}
\begin{document}
\maketitle
\begin{abstract}
Counterfactual explanations have emerged as a popular tool to explain
predictions made by opaque machine learning models: they explain how
factual inputs need to change in order for some fitted model to produce
some desired output. Much existing research has focused on identifying
explanations that are not only valid but also deemed plausible and
desirable with respect to the underlying data and stakeholder
requirements. Recent work has shown that under this premise, the task of
learning plausible explanations is effectively reassigned from the model
itself to the (post-hoc) counterfactual explainer. Building on that
work, we propose a novel model objective that leverages counterfactuals
during the training phase (ad-hoc) in order to minimize the divergence
between learned representations and plausible explanations. Through
extensive experiments, we demonstrate that our proposed methodology
facilitates training models that inherently deliver plausible
explanations while maintaining high predictive performance.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
Counterfactual Explanations \sep Explainable AI \sep 
Representation Learning

@@

Part of the problem is that high degrees of freedom provide room for
many solutions that are locally optimal with respect to narrow
objectives (\citeproc{ref-wilson2020case}{Wilson 2020}).\footnote{For
  clarity: we follow standard ML convention in using ``degrees of
  freedom'' to refer to the number of parameters estimated from data.}
Based purely on predictive performance, these solutions may seem to
provide compelling explanations for the data, when in fact they are
based on purely associative, semantically meaningless patterns. This
poses two related challenges: firstly, it makes these models inherently
opaque, since humans cannot simply interpret what type of explanation
the complex learned representations correspond to; secondly, even if we
could resolve the first challenge, it is not obvious how to mitigate
models from learning representations that correspond to meaningless and
implausible explanations.

@@

Counterfactual training combines ideas from adversarial training,
energy-based modelling and counterfactuals explanations with the
explicit objective of aligning representations with plausible
explanations that comply with user requirements. In the context of CE,
plausibility has broadly been defined as the degree to which
counterfactuals comply with the underlying data generating process
(\citeproc{ref-poyiadzi2020face}{Poyiadzi et al. 2020};
\citeproc{ref-guidotti2022counterfactual}{Guidotti 2022};
\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}). Plausibility
is a necessary but insufficient condition for using CE to provide
algorithmic recourse (AR) to individuals affected by opaque models in
practice. This is because for recourse recommendations to be
\textbf{actionable}, they need to not only result in plausible
counterfactuals but also be attainable. A plausible CE for a rejected
20-year-old loan applicant, for example, might reveal that their
application would have been accepted, if only they were 20 years older.
Ignoring all other features, this complies with the definition of
plausibility if 40-year-old individuals are in fact more credit-worthy
on average than young adults. But of course this CE does not qualify for
providing actionable recourse to the applicant since \emph{age} is not a
(directly) mutable feature. For our intents and purposes, counterfactual training
aims at improving model explainability by aligning models with
counterfactuals that meet both desiderata, plausibility and
actionability. Formally, we define explainability as follows:

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  (Plausibility)
  \(\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1\)
  where \(A\) is some small region around \(\mathbf{x}^{\prime}\).
\item
  (Actionability) Permutations \(\Delta\) are subject to some actionability
  constraints.
\end{enumerate}

@@

Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote a counterfactual
explanation generated through gradient descent over \(T\) iterations as
initially proposed by Wachter, Mittelstadt, and Russell
(\citeproc{ref-wachter2017counterfactual}{2017}). For our purposes, we
let \(T\) vary and consider the counterfactual search as converged as
soon as the predicted probability for the target class has reached a
pre-determined threshold, \(\tau\):
\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\), where \(\mathcal{S}\) is the softmax function.\footnote{For
  detailed background information on gradient-based counterfactual
  search and convergence see Section~\ref{sec-app-ce}.}

@@

Intuitively, the gradient of Equation~\ref{eq-div} decreases the energy
of observed training samples (positive samples) while at same time
increasing the energy of counterfactuals (negative samples)
(\citeproc{ref-du2019implicit}{Du and Mordatch 2020}). As the generated
counterfactuals get more plausible (Definition~\ref{def-explainability})
over the course of training, these two opposing effects gradually balance
each out (\citeproc{ref-lippe2024uvadlc}{Lippe 2024}).

@@

Under this choice, we consider nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\) as adversarial examples as long as the
magnitude of the perturbation to any individual feature is at most
\(\varepsilon\). This is closely aligned with Szegedy et al.
(\citeproc{ref-szegedy2013intriguing}{2013}), who define an adversarial
attack as an ``imperceptible non-random perturbation''. Thus, we choose
to work with a different distinction between CE and AE than Freiesleben
(\citeproc{ref-freiesleben2022intriguing}{2022}), who considers
misclassification as the key distinguishing feature of AE. One of the
key observations in this work is that we can leverage counterfactual
explanations during training and get adversarial examples essentially
for free.
